---
title: "Mark and Jiawen Climate Project"
author: "Carl Boettiger & Dana Seidel & Mark Sun & Jiawen Tang"
output: github_document
---

# Unit I: Climate Change Module

## Warm-up exercise: Examining CO2 trends in R

-   Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
-   Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

```{r message=FALSE}
library(tidyverse)
```

```{r message=FALSE}

co2 <- 
read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date", "monthly_average",
                                "interpolated", "days","std", ""),
                  na = c("-1", "-9.99","-0.99"))
co2
```

```{r}
ggplot(co2, aes(x = decimal_date)) + 
  geom_line(aes(y=monthly_average),col="blue")+
  geom_line(aes(y=interpolated),col="red")

```

Which months are the CO2 values at the maximum? Minimum? Why is this? May is the maximum and October is the minimum; since 90% of the FF emission in northern hemisphere and photosynthesis will help reduce. What rolling average is used in computing the "trend" line? How does the trend depend on the rolling average?

------------------------------------------------------------------------

# Exercise I: Temperature Data

Each of the last years has consecutively set new records on global climate. In this section we will analyze global mean temperature data.

Data from: <http://climate.nasa.gov/vital-signs/global-temperature>

## Question 1:

Describe the data set to the best of your ability given the documentation provided. Describe what kind of column each data contains and what units it is measured in. Then address our three key questions in understanding this data:

-   How are the measurements made? What is the associated measurement ?
-   What is the resolution of the data?
-   Are their missing values? How should they be handled?

The data source is NASA/GISS. While the article doesn't articulate how the measurements are made,it is likely they study ice cores to construct past global mean temperature.The associated measurement may be inherent difficulty to make exact predictions. The resolution of the data, which refers to "the period of time represented by each data point that is collected for the report", is a year, since every data point corresponds to each year's annual mean temp. The data set is complete (no missing values), if there are, they may be represented with "NA" and wiill not be considered in analysis.

## Question 2:

Construct the necessary R code to import and prepare for manipulation the following data set: <http://climate.nasa.gov/system/internal_resources/details/original/647_Global_Temperature_Data_File.txt>

```{r}
climate <- read_table("http://climate.nasa.gov/system/internal_resources/details/original/647_Global_Temperature_Data_File.txt",
           col_names = c("year", "annual_average", "five_year_average"),
           na="NA",
           col_types = ('ddd'),
           skip = 5
           )
climate
```

## Question 3:

Plot the trend in global mean temperatures over time. Describe what you see in the plot and how you interpret the patterns you observe.

```{r}

ggplot(climate, aes(x = year, y = annual_average)) + geom_line()
```

There is a annual variation in the temperature mean, but the trend is increasing after 1920 and with a stronger signal after 1960. It indicates the temperature is increase compare to average.

## Question 4: Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously. This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone. (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)).

By examining the data here, what evidence do you find or not find for such a pause? Present an analysis of this data (using the tools & methods we have covered in Foundation course so far) to argue your case.\
What additional analyses or data sources would better help you refine your arguments?

By looking at the data, I did not find evidence in support of such a hiatus around 1998. The graph above showing global mean temperatures over time depicts a rather consistent, steady raise in global mean temperature since 1998.There is not a pause or decrease in the trend. Some additional analysis that may help could be determining the rate of global temperature change by performing regression analysis.

## Question 5: Rolling averages

-   What is the meaning of "5 year average" vs "annual average"? 5 year average is the rolling average that smooth the annual variation signal in the data, and compare to annual average, it shows a more smooth trend of increase in temperature.
-   Construct 5 year averages from the annual data. Construct 10 & 20-year averages.
-   Plot the different averages and describe what differences you see and why.

```{r}
ggplot(climate,aes(x= year, y = five_year_average)) + geom_line()
#ggplot(co2, aes(x = decimal_date, y = monthly_average)) + geom_line() 

library(dplyr)
library(zoo)

#calculate 10-year & 20-year rolling average
climate = climate %>%
  mutate(ten_year_avg = rollmean(climate$annual_average, k=10, fill=NA, align='right'),
         twenty_year_avg = rollmean(climate$annual_average, k=20, fill=NA, align='right'));
climate
```

```{r}
ggplot(climate,aes(x= year)) + 
  geom_line(aes(y = ten_year_avg, color = 'ten')) + 
  geom_line(aes(y = twenty_year_avg, color = 'twenty')) +
  labs(y = "Temperature Change (C)", color = NULL) +
  scale_color_manual(values = c(ten = "blue", twenty = "green"),
                      labels = c(ten = "ten_year_avg", twenty = "twenty_year_avg"),
                      limits = c("ten", "twenty"))
```

The data looks more flat as the average year increases, becoming more pronounced and losing some of the temperature fluctuations caused by special events.

# Exercise II: Melting Ice Sheets?

-   Data description: <http://climate.nasa.gov/vital-signs/land-ice/>
-   Raw data file: <http://climate.nasa.gov/system/internal_resources/details/original/499_GRN_ANT_mass_changes.csv>

## Question 1:

-   Describe the data set: what are the columns and units? Where do the numbers come from?

    The columns are time in decimal years, Greenland and Antarctica ice sheet mass in Giga tonnes. The data are collected by NASA GRACE and GRACE Follow-On satellites, which are gravitational satellites that measure land change by change of gravitational force.

-   What is the measurement? Resolution of the data? Interpretation of missing values?

    The GRACE mission ended in June 2017. The GRACE Follow-On mission began collecting data in June 2018, with some lost data in the middle. The record includes new data processing methods and is constantly updated as more data comes in, with a delay of up to two months. Ice sheet data are measured about 12 times a year.

## Question 2:

Construct the necessary R code to import this data set as a tidy `Table` object.

```{r}
IceMass<-read_csv("http://climate.nasa.gov/system/internal_resources/details/original/499_GRN_ANT_mass_changes.csv",
              col_names = c("year","greenland","antarctica"),
              col_types = ('ddd'),
              skip = 10)
IceMass

```

## Question 3:

Plot the data and describe the trends you observe.

```{r}
ggplot(IceMass,aes(x= year)) + 
  geom_line(aes(y = greenland, colour = 'Greenland' )) + 
  geom_line(aes(y = antarctica, colour = 'Antarctica')) +
  ggtitle('Icesheet mass change over time') +
  labs(y = "Change of ice mass (Gt)", color = NULL)
```

Both ice sheets showed a clear declining trend and the presence of seasons was evident. Greenland, which is located in the Arctic, was affected by positive feedback more, and the weight of the ice sheet dropped faster.

# Exercise III: Rising Sea Levels?

-   Data description: <http://climate.nasa.gov/vital-signs/sea-level/>
-   Raw data file: <http://climate.nasa.gov/system/internal_resources/details/original/121_Global_Sea_Level_Data_File.txt>

## Question 1:

-   Describe the data set: What are the columns and units?

    There are 12 columns in the data set and we are only going to present 3 of them to make the table clean and useful. The 12 columns are: HDR 1 altimeter type 0=dual-frequency 999=single frequency (ie Poseidon-1) HDR 2 merged file cycle \# HDR 3 year+fraction of year (mid-cycle) HDR 4 number of observations HDR 5 number of weighted observations HDR 6 GMSL (Global Isostatic Adjustment (GIA) not applied) variation (mm) with respect to TOPEX collinear mean reference HDR 7 standard deviation of GMSL (GIA not applied) variation estimate (mm) HDR 8 smoothed (60-day Gaussian type filter) GMSL (GIA not applied) variation (mm)\
    HDR 9 GMSL (Global Isostatic Adjustment (GIA) applied) variation (mm) with respect to TOPEX collinear mean reference HDR 10 standard deviation of GMSL (GIA applied) variation estimate (mm) HDR 11 smoothed (60-day Gaussian type filter) GMSL (GIA applied) variation (mm) HDR 12 smoothed (60-day Gaussian type filter) GMSL (GIA applied) variation (mm); annual and semi-annual signal removed

-   Where do these data come from?

    The calculations were performed by NASA's Goddard Space Flight Center with support from NASA's Measurement Program. GMSL is generated using integrated multi-mission ocean altimeter data. It combines sea surface heights from TOPEX/Poseidon, Jason-1, and OSTM/Jason-2 into a common ground reference frame.

-   What is the in-measurement? Resolution of the data? Interpretation of missing values?

    The data applies and places all inter-mission bias, range, and geophysical corrections in a georeference orbit. This creates a consistent data record throughout time, regardless of the instrument used.

## Question 2:

Construct the necessary R code to import this data set as a tidy `Table` object.

```{r}
SeaLevel<-read_table("http://climate.nasa.gov/system/internal_resources/details/original/121_Global_Sea_Level_Data_File.txt",
                     skip=45,
                     col_names = c("a","b","year","d","e","f","g","h","GMSL","j","k","Smoothed_60_days"),
                     na="99900")

SeaLevel<- SeaLevel[ ,c('year', 'GMSL', 'Smoothed_60_days')]
SeaLevel

```

## Question 3:

Plot the data and describe the trends you observe.

```{r}
ggplot(SeaLevel,aes(x= year)) + 
  geom_line(aes(y = GMSL, colour = 'GMSL' )) + 
  geom_line(aes(y = Smoothed_60_days, colour = 'Smoothed_60')) +
  ggtitle('Global Mean Sea Level change over time') +
  labs(y = "Change of sea level(mm)", color = NULL)
```

The trend of sea level rise is very clear, and the seasonal and meteorological effects are very strong, as can be seen by comparing the rough curve. However, it has risen by about 10cm in 30 years, which is already a very obvious upward trend.

# Exercise IV: Arctic Sea Ice?

-   <http://nsidc.org/data/G02135>
-   <ftp://sidads.colorado.edu/DATASETS/NOAA/G02135/north/daily/data/N_seaice_extent_daily_v3.0.csv>

## Question 1:

-   Describe the data set: what are the columns and units?

There are 6 columns in the original file. Three of them are the time (in year, month, and day of each recording. One column represents the quantity (extent) of sea ice in 10\^6 sq km. The other 2 columns indicates missing data and source of data. We expurgated the data and left only 2 columns: the time of recording and the extent of sea ice at that tiume.

-   Where do these data come from?

The data come from NSIDC (National Snow and Ice Data Center)

-   What is the uncertainty in measurement? Resolution of the data? Interpretation of missing values?

The uncertainty in measurement may arise from tiny inaccuracy of data measurment tools. Besides, the resort to time-series analysis may involve small extent of interpolation, so the measurement may not be exactly accurate and certain.The resolution of the data is 2 days, as each recording time is separated by 2 days.There is no missing values as they are all left as 0. In other words, this data sheet is consistent and complete.

## Question 2:

Construct the necessary R code to import this data set as a tidy `Table` object.

```{r}
ArcticIce<-read_csv("ftp://sidads.colorado.edu/DATASETS/NOAA/G02135/north/daily/data/N_seaice_extent_daily_v3.0.csv",
              #col_names = c("year","greenland","antarctica"),
              col_types = ('ddddd'),
              skip = 0 )
ArcticIce$Date<-as.Date(with(ArcticIce,paste(Year,Month,Day,sep="-")),"%Y-%m-%d")
ArcticIce <- ArcticIce [ ,c('Date', 'Extent') ]
ArcticIce
```

## Question 3:

Plot the data and describe the trends you observe.

As shown in the plot, we observe a constant, gradually declining trend of Sea Ice through time, from 1980 to 2020. There are also significant seasonal fluctuations of Sea Ice each year. Overall, it's clear that more and more Sea Ice is disappearing.

```{r}
ggplot(ArcticIce,aes(x= Date)) + 
  geom_line(aes(y = Extent, colour = 'Sea Ice' )) + 
  #geom_line(aes(y = Smoothed_60_days, colour = 'Smoothed_60')) +
  ggtitle(' Sea Ice Index ') +
  labs(y = "Sea Ice (10^6 sq km)", color = NULL) +
  geom_smooth(aes(y = Extent, colour = 'trend'), method=lm)
```

# Exercise V: Longer term trends in CO2 Records

The data we analyzed in the unit introduction included CO2 records dating back only as far as the measurements at the Manua Loa observatory. To put these values into geological perspective requires looking back much farther than humans have been monitoring atmosopheric CO2 levels. To do this, we need another approach.

[Ice core data](http://cdiac.ornl.gov/trends/co2/ice_core_co2.html):

Vostok Core, back to 400,000 yrs before present day

-   Description of data set: <https://www.antarcticglaciers.org/glaciers-and-climate/ice-cores/ice-core-basics/#:~:text=420%2C000%20years%20of%20ice%20core%20data%20from%20Vostok%2C%20Antarctica%20research%20station.>
-   Data source: <https://d32ogoqmya1dw8.cloudfront.net/files/NAGTWorkshops/environmental/workshop12/vostok_ice_core_data.xls>

```{r}
library(readxl)
url <- "https://d32ogoqmya1dw8.cloudfront.net/files/NAGTWorkshops/environmental/workshop12/vostok_ice_core_data.xls"
destfile <- "vostok_ice_core_data.xls"
curl::curl_download(url, destfile)
sheet_index <- 2
Icecore_co2 <- read_excel(destfile, sheet = sheet_index,
                          col_names = c("time","co2","resolution"),
                          skip = 1)
Icecore_co2

```

## Questions / Tasks:

-   Describe the data set: what are the columns and units? Where do the numbers come from?

The data set we found contains three columns: time (representing years before present) ,co2 concentration, and resolution of the data. For instance, the time 0 mean present, and time 2342 means 2342 years before now.

-   What is the uncertainty in measurment? Resolution of the data? Interpretation of missing values?

Uncertainty may arise from tiny inaccuracy when interpolating co2 level long time ago. The resolution of the data varies from hundreds to thousands, usually aroune 1000-2000 years.There is no missing value in this data set.

-   Read in and prepare data for analysis.

```{r}
climate <-ggplot(Icecore_co2,aes(x=time)) + 
  geom_line(aes(y=co2))
climate
```

-   Reverse the ordering to create a chronological record.

```{r}
library(dplyr)
Icecore_co2$time <- 414085-Icecore_co2$time
Icecore_co2 <- Icecore_co2[ ,c('time', 'co2')]
Icecore_co2
```

-   Plot data

```{r}
Icecore_co2
ggplot(Icecore_co2,aes(x=time))+
  geom_line(aes(y= co2))+
  labs(x = "years from 414085 years BP to present", title = "Past 400000 years trends in CO2 Records") +
  theme(axis.text.x = element_blank())
  
```

-   Join this series to Loa data

```{r}
co2_2009 <- co2[ ,c('decimal_date', 'interpolated')]
co2_2009 <- co2_2009[co2_2009$decimal_date>2009,]
co2_2009$decimal_date <- co2_2009$decimal_date+414085-2009
colnames(co2_2009) <- c("time", "co2")
co2_2009


full_co2<-rbind(Icecore_co2,co2_2009)
full_co2<-full_co2 %>% arrange((time))
# Reorder the table by the "Score" column in decreasing order
full_co2

```
-   Plot joined data

```{r}
#Icecore_co2
ggplot(full_co2,aes(x=time))+
  geom_line(aes(y = co2))+
  labs(x = "years from 414099 years BP to present", title = "Past 400000 years trends in CO2 Records") +
  theme(axis.text.x = element_blank())
```

-   Describe your conclusions
Based on the Ice core co2 data, we show constant, moderate natural variations in co2 level since 400000 years ago, and the range is around 175 to 300 ppm. However, since Industrial revolution which happened "recently" in the long earth history, co2 level rises at an unprecedently considerable rate, as manifested by the almost vertical straight line.This recent surge in co2 is thus unlikely caused solely by nature: human activities have played a great role.
